---
title: "JumpForests: An R package for random forests for jump processes"
output: github_document
author: "Martin Bladt and Rasmus Frigaard Lemvig"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Plan:
- The simple time-homogeneous Markov model with one covariate
- A simulated very simple life insurance dataset with Poisson regression as benchmark, tune for the best splitting rule
- Analysis of a real dataset (bmt?)

This vignette highlights some simple applications of the *JumpForests* package. The package implements trees and random forests for regression, classification, survival and general multi-state models.
As the latter is a novel contribution, we focus on this application in this vignette. To install the package, simply run the following command in R:

```{r install, eval=FALSE}
remotes::install_github("martinbladt/JumpForests", upgrade = "never")
```

We illustrate the use of the package in three cases, a simple time-homogeneous Markov model with one covariate, a simulated very simple life insurance dataset with Poisson regression as benchmark, and an analysis of a real dataset.

## A Markov model with independent censoring and covariates

We consider the following simple time-homogeneous Markov model with one covariate which was also illustrated in the vignette of the *AalenJohansen* package [here](https://cran.r-project.org/web/packages/AalenJohansen/vignettes/AalenJohansen-vignette.html).

```{=latex}
\begin{align*}
\frac{\mathrm{d}\Lambda(t|x)}{\mathrm{d}t}=\lambda(t|x)=\frac{1}{1+x\cdot t}
\begin{pmatrix}
-2 & 1& 1 \\
2 & -3 & 1 \\
0 & 0 & 0
\end{pmatrix}\!.
\end{align*}
```

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
